---
title: "Problem Set 3"
author: "Patrick Altmeyer"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  bookdown::html_document2:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: false
bibliography: bib.bib
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(data.table)
library(expm)
library(igraph)
helper_functions <- list.files("R")
lapply(
  helper_functions,
  function(i) {
    source(file = file.path("R",i))
  }
)
```

# Exercise 1 - Erdos-Renyi

## a)

Using the model $G(n,p)$, create an undirected Erdos-Renyi network with 100 nodes
and probability of each link equal to 0.02. For each of these networks:

<hr>

We can use a simple helper function to randomly generate a $G(n,p)$ model in R:

```{r, code=readLines("R/utils.erdos_renyi.R"),echo=TRUE,eval=FALSE}
```

Let us set a seed and use the function to generate a random network:

```{r}
set.seed(42)
G <- erdos_renyi()
g <- graph_from_adjacency_matrix(G, mode = "undirected")
```

### i)

use the Kamada-Kawai model to represent the network,

<hr>

```{r}
coords <- igraph::layout.kamada.kawai(g)
plot.igraph(g, layout=coords)
```

### ii)

find its degree distribution; what is the degree of the more connected node? And what is the degree of the less connected one?

<hr>

```{r}
d <- diag(crossprod(G))
```


The degree distribution can be computed through the following simple helper function:

```{r, code=readLines("R/utils.degree_dist.R")}
```

For the randomly generated network Figure (\@ref(fig:deg-dist)) shows the corresponding degree distribution. Evidently, the more connected node is of degree of $d_{max}=`r max(d)`$ and the less connected node is of degree $d_{min}=`r min(d)`$.

```{r deg-dist, fig.cap="Degree distribution of randomly generated network."}
deg_dist <- degree_dist(G)
```

### iii)

is there one or more than one connected components?

<hr>

```{r}
comp <- components(g)
n_comp <- comp$no
```

There are `r n_comp` components: `r sum(comp$csize>1)` of size greater than one and `r sum(comp$csize==1)` that count just one node:

```{r}
comp_greater_one <- unique(comp$membership[comp$membership %in% which(comp$csize>1)])
mark_groups <- lapply(
  comp_greater_one,
  function(i) {
    which(comp$membership==i)
  }
) 
plot.igraph(g, layout=coords, mark.groups = mark_groups)
```

### iv)

what is the average distance? And the diameter? Interpret the numbers you
obtain.

<hr>



### v)

what is the number of type 300 triads (according to M-A-N labeling, as described
by Holland, Leinhardt and Davis) in the triad census report?

<hr>

### vi)

find the local/individual clustering coefficient distribution. Compute the average
clustering coefficient.

<hr>

## b)

Repeat the exercise but with the probability of each link equal to 0.08.

<hr>

## Exercise 2 - Scale-Free (Barabasi-Albert’s model)

Now create a directed network with 100 nodes using a linear preferential attachment model: in each period one new node enters in the network and creates one link towards another one; initially there is just one single isolated node. There is another parameter, the zero appeal. Explain first of all what is the definition of this parameter.

### i)

create first a network in which the zero appeal is equal to 0. What happens in that case? Why?

<hr>

### ii) 

create now another network in which the zero appeal is equal to 1.

<hr>

### iii)

plot the network using the Kamada-Kawai model.

<hr>

### iv)

study the in-degree distribution.

<hr>

### v)

transform the network you have obtained into an undirected one. What is the number of components in that network? What is the average distance? What is the diameter?

<hr>

### vi)

explore what happens if we increase the zero appeal parameter.

<hr>
