---
title: "Informal collaboration in financial economics"
subtitle: "Report"
author: "Patrick Altmeyer"
date: "`r format(Sys.Date(), '%d %B, %Y')`"
output: 
  bookdown::html_document2:
    code_folding: show
    toc: true
    toc_float: true
    number_sections: false
bibliography: bib.bib
---

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
library(data.table)
library(expm)
library(igraph)
library(reticulate)
helper_functions <- list.files("R")
lapply(
  helper_functions,
  function(i) {
    source(file = file.path("R",i))
  }
)
```

- is the question relevant?

- are network interactions relevant in that setting? Why?

- what is/are the main takeaway(s) from the paper?

- is there anything you particularly like or dislike from the methodology/analysis? Would you suggest a different/complementary approach?

- do you think there are open issues/questions that could be considered?

- is there anything that could extrapolate to other situations as well?

- have you found in the paper what you were expecting? If not, how does it differ from your expectations?

# Motivation {#mot}

- Formal collaboration in research has been show to improve productivity
- Less research has been done on the impact of informal collaboration on research outcomes: conferences, seminars, informal reviews from colleagues etc.
- "Much if not most of the debate and discussion about economic ideas take place at the pre-working paper, workshop and working paper stages."

# Contributions

- The network of informal collaborations may be useful as explanatory variable when studying e.g. the determi- nants of scientific productivity. Moreover, it may serve as an interesting dependent variable in studies exploring the relationship of individual characteristics and social capital.

# Data {#data}

- novel dataset on informal collaboration in Financial Economics obtained from the acknowledgement sections of over 5, 000 published research papers

- From the analysis we omit research assistants, editorial support and non-academic commenters (such as industry professionals or central bankers) if they are acknowledged as such. 

**Researcher characteristics**

- Combining both the number of citations and publications we measure prolificness using the Euclidean index of citations.14 

- Next we define experience as the number of years between the first publication and the year of publication of the paper

Gender is estimated from first names. 



Data has been made publically available...

```{python, warnings=FALSE}
from json import loads
from urllib.request import urlopen
COFE = 'https://raw.githubusercontent.com/Michael-E-Rose/CoFE/master/acks_min.json'
acks = loads(urlopen(COFE).read().decode("utf-8"))['data']
```

```{python, warnings=FALSE}
from itertools import product
import networkx as nx
import numpy as np

G = nx.Graph()
for item in acks:
    # Authors
    auths = [a['label'] for a in item['authors']]
    G.add_nodes_from(auths)
    # Commenters
    coms = [c['label'] for c in item.get('com', [])]
    G.add_nodes_from(coms)
    # Links
    links = list(product(coms, auths))
    G.add_edges_from(links)
    
A = nx.adjacency_matrix(G)
```

```{r}
g <- graph_from_adjacency_matrix(py$A)
```

- Threads to validity: While strategic acknowledging is discouraged and can be costly, it remains the main threat to reliability measuring informal collaboration with acknowledgements data.

# Methodology {#meth}

-  In the network of infor- mal collaboration, two researchers are connected with a weighted directed link whenever one acknowledges the other on a published paper in our dataset. 

Use different centrality measures: degree, eigen and betweenness. Focus is limited to "giant" component since centrality measures are not comparable across components (**DOUBLE CHECK. POTENTIAL SHORTFALL?**).

- the effort brought forward in equilibrium corresponds to someone’s eigenvector centrality.

- Since eigenvector centrality focuses on connectivity and influence only, but remains silent about the importance of a researcher for knowledge flows, we also study betweenness centrality

- More senior researchers are less influential in the network (lower eigenvector centrality), while more prolific researchers are more likely to connect otherwise disparate research communities (higher betweenness centrality). 

# Findings {#findings}

- researcher’s position in this network is predictive of her future productivity and the scholarly impact of the papers she comments on

- We find positive but rather weak correlations (around 0.5) between scientific productivity measures and the number of acknowledged contributions.

- Female researchers are ackowledged considerably less

- find centralities in the network of informal collaboration to have higher explanatory power than centralities in the co-author networks.

